---
title: "Ongoing containment as a policy objective"
author: Steven Riley
date: late 2023
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ongoing containment as a policy objective}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Background

This analysis was initially developed between mid-February and early morning of March 10th 2020. It was used during February to give early estimates of the possible impact of school closures then updated to describe what failed mitigation would look like. The original form of the script used for the paper on ongoing containment is presevered in the R file.

Next steps

* add the processed excess mortality as a data object to the package

## Load functions and population data

First clear memory

```{r }
rm(list=ls(all=TRUE))
```

Load functions to run the hybrid model from local packages, from github or directly from local copies
if that is how the code has been distributed.

```{r }
## source("~/git/idd/R/cov_hybrid.r")
## library(devtools)
library(dplyr)
```

Install SR's idd package either from github or locally

```{r }
## install_github("c97sr/idd")
## devtools::uninstall()
## devtools::load_all()
## library("idd")
```

Setup a population of size 66 million with a small timestep, and assume
that there were 6000 infectious individuals at time 0. There are four groups in the model which would normally represent age groups. However, they are all assumed to be eqactly equal.

```{r }
dt <- 0.1
pop <- 66000000*c(0.25,0.25,0.25,0.25)
seed <- rep(6000/length(pop),length(pop))
xvals <- seq(0,540,dt)
```

Make an initial run for 2 years with an R0 of 2.0 in a UK-like
population, with a trickle seed of 10 cases per week and nop
reactive behaviour change.

```{r }
y1 <- idd::cov_hybrid(
    vecN=pop,
    vecI0=seed,
    R0=2.0,
    Rp=1.4,
    Rover=0.95,
    trig_pres=99999999,
    icu_cap=4357/53000000,
    sevBchange=FALSE,
    vecPS=rep(0.05,length(pop)),
    trickle=10,
    vecTcalc=xvals)
```

Then run with a reactive drop when the virus is known to be present

```{r }
y2 <- idd::cov_hybrid(
    vecN=pop,
    vecI0=seed,
    R0=2.0,
    Rp=1.4,
    Rover=0.95,
    trig_pres=5,
    icu_cap=4357/53000000,
    sevBchange=FALSE,
    vecPS=rep(0.05,length(pop)),
    trickle=10,
    vecTcalc=xvals)
```

Then run with a reactive drop when the virus is known to be present

```{r }
y3 <- idd::cov_hybrid(
    vecN=pop,
    vecI0=seed,
    R0=2.0,
    Rp=0.95,
    Rover=0.95,
    trig_pres=5,
    icu_cap=4357/53000000,
    sevBchange=FALSE,
    vecPS=rep(0.05,length(pop)),
    trickle=10,
    vecTcalc=xvals)
```

Then run with a reactive drop when the virus is known to be present

```{r }
y4 <- idd::cov_hybrid(
    R0=2.0,
    vecN=pop,
    vecI0=seed,
    Rp=1.4,
    Rover=0.99,
    trig_pres=5,
    icu_cap=4357/53000000,
    sevBchange=TRUE,
    vecPS=rep(0.05,length(pop)),
    trickle=10,
    vecTcalc=xvals)
```

Then run with a reactive drop when the virus is known to be present

```{r }
y4A <- idd::cov_hybrid(
    R0=2.0,
    vecN=pop,
    vecI0=seed,
    Rp=1.4,
    Rover=0.99,
    trig_pres=5,
    icu_cap=2*4357/53000000,
    sevBchange=TRUE,
    vecPS=rep(0.05,length(pop)),
    trickle=10,
    vecTcalc=xvals)
y4B <- idd::cov_hybrid(
    R0=2.0,
    vecN=pop,
    vecI0=seed,
    Rp=1.4,
    Rover=1.1,
    trig_pres=5,
    icu_cap=4357/53000000,
    sevBchange=TRUE,
    vecPS=rep(0.05,length(pop)),
    trickle=10,
    vecTcalc=xvals)
```

Plot the baselines

```{r }
par(mfrow=c(1,2))
for (s in c("","y")) {
    plot(y1$t/30,rowSums(y1$inf[,,1]/dt+1),
         log=s,col="red",type="l",lwd=2,xlim=c(0,18),
         xlab="Month",ylab="Daily incidence (+1)")
    points(y2$t/30,rowSums(y2$inf[,,1]/dt+1),col="blue",type="l",lwd=2)
    points(y3$t/30,rowSums(y3$inf[,,1]/dt+1),col="green",type="l",lwd=2)
    points(y4$t/30,rowSums(y4$inf[,,1]/dt+1),col="cyan",type="l",lwd=2)
}
```

Plot the sensitvities

```{r }
par(mfrow=c(1,2))
for (s in c("","y")) {
    plot(y1$t/30,rowSums(y1$inf[,,1]/dt+1),
         log=s,col="gray",type="l",lwd=2,xlim=c(0,18),
         xlab="Month",ylab="Daily incidence (+1)")
    points(y2$t/30,rowSums(y2$inf[,,1]/dt+1),col="gray",type="l",lwd=2)
    points(y3$t/30,rowSums(y3$inf[,,1]/dt+1),col="gray",type="l",lwd=2)
    points(y4$t/30,rowSums(y4$inf[,,1]/dt+1),col="gray",type="l",lwd=2)
    points(y4A$t/30,rowSums(y4A$inf[,,1]/dt+1),col="orange",type="l",lwd=2)
    points(y4B$t/30,rowSums(y4B$inf[,,1]/dt+1),col="magenta",type="l",lwd=2)
}
```

Define a function to then output the total amount of infection

```{r }
tabdisp <- function(x)(formatC(signif(x,3), format="f", big.mark=",", digits=0))
tabdisp(sum(y1$inf[,,1]))
tabdisp(sum(y2$inf[,,1]))
tabdisp(sum(y3$inf[,,1]))
tabdisp(sum(y4$inf[,,1]))
tabdisp(sum(y4A$inf[,,1]))
tabdisp(sum(y4B$inf[,,1]))
```

Output the total number of deaths

```{r }
tabdisp(y1$sevIn*0.2 + y1$sevOut)
tabdisp(y2$sevIn*0.2 + y2$sevOut)
tabdisp(y3$sevIn*0.2 + y3$sevOut)
tabdisp(y4$sevIn*0.2 + y4$sevOut)
tabdisp(y4A$sevIn*0.2 + y4A$sevOut)
tabdisp(y4B$sevIn*0.2 + y4B$sevOut)
```

Plot cumulative deaths for different scenarios
Plot

```{r }
par(mfrow=c(1,1))
plot(0:1,type="n",ylim=c(0,4000),xlim=c(0,600/7),
     ylab="Cum COVID-19 Deaths",xlab="Time")
y_scen1 <- (rowSums(0.2*y1$csevtreat[,,1])+rowSums(y1$csevuntreat[,,1]))/sum(pop)*100000
points(y1$t/7,y_scen1,type="l")
y_scen2 <- (rowSums(0.2*y2$csevtreat[,,1])+rowSums(y2$csevuntreat[,,1]))/sum(pop)*100000
points(y2$t/7,y_scen2,type="l")
y_scen4 <- (rowSums(0.2*y4$csevtreat[,,1])+rowSums(y4$csevuntreat[,,1]))/sum(pop)*100000
points(y4$t/7,y_scen4,type="l")
y_scen4a <- (rowSums(0.2*y4$csevtreat[,,1])+0.6*rowSums(y4$csevuntreat[,,1]))/sum(pop)*100000
points(y4$t/7,y_scen4a,type="l")
```

Output the proportion of herd immunity acheived at 18 months

```{r }
round(sum(y1$inf[,,1])/(sum(pop)/2)*100)
round(sum(y2$inf[,,1])/(sum(pop)/2)*100)
round(sum(y3$inf[,,1])/(sum(pop)/2)*100)
round(sum(y4$inf[,,1])/(sum(pop)/2)*100)
round(sum(y4A$inf[,,1])/(sum(pop)/2)*100)
round(sum(y4B$inf[,,1])/(sum(pop)/2)*100)
```

Work to load up excess mortality data. This will go futher up in the final
version of this

```{r }
```

These bits need to be coded as a data object

```{r }
if (FALSE) {
  ## tmpdir <- "C:/tmp/"
  localfn <- paste0(tmpdir,"excess_death.csv")
  localfn_w <- paste0(tmpdir,"excess_death_econ.csv")
  ## webfn <- "https://raw.githubusercontent.com/owid/covid-19-data/4132d2c726fa406d5ff934b4f6411f463992a07d/public/data/excess_mortality/excess_mortality.csv"
  ## x <- read.csv(webfn)
  ## write.csv(x,file=localfn,row.names = FALSE)
  ## wwebfn <- "https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/excess_mortality/excess_mortality_economist_estimates.csv"
  ## w <- read.csv(wwebfn)
  ## write.csv(w,file=localfn_w,row.names=FALSE)
  
  y <- read.csv(localfn)
  z <- read.csv(localfn_w)
  names(z)
  dim(z)
  
  length(table(z$country))
  
  ## all(dim(x) == dim(y))
  ## all(names(x) == names(y))
  df <- y %>%
    dplyr::filter(time_unit=="weekly", date < as.Date("2021-08-31")) %>%
    dplyr::select("location", "cum_excess_per_million_proj_all_ages","date") %>%
    dplyr::rename("cumexcess" = "cum_excess_per_million_proj_all_ages",
                  "country"="location") %>%
    tidyr::spread(country,cumexcess)
  dim(df)
  names(df)
  
  df2 <- z %>%
    dplyr::filter(date < as.Date("2021-08-31")) %>%
    dplyr::select("date","country",
                  "cumulative_estimated_daily_excess_deaths_per_100k") %>%
    dplyr::rename("cumexcess" =
                    "cumulative_estimated_daily_excess_deaths_per_100k") %>%
    tidyr::spread(country,cumexcess)
  
  dim(df2)
  colnames <- names(df2)
  windex <- which(colnames=="World")
  dindex <- which(colnames=="date")
  countries <- colnames[-c(windex,dindex)]
  ncountries <- length(countries)
  
  plot_cum <- function(ymin=-10,ymax=4000, weekmax=75,incmodels=TRUE) {
    plot(0:1,type="n",ylim=c(ymin,ymax),xlim=c(0,weekmax),
         ylab="Excess mortality per 100k",xlab="Week 2020")
    for (c in countries) {
      points(df2[1:weekmax,c],type="l",col="grey",lwd=0.5)
    }
    points(df2$`United Kingdom`[1:weekmax],type="l",col="blue",lwd=3)
    points(df2$Australia[1:weekmax],type="l",col="red",lwd=3)
    points(df2$China[1:weekmax],type="l",col="green",lwd=3)
    points(df2$`United States`[1:weekmax],type="l",col="cyan",lwd=3)
    points(df2$Norway[1:weekmax],type="l",col="magenta",lwd=3)
    ## points(df2$Taiwan[1:weekmax],type="l",col="orange",lwd=3)
    if (incmodels) {
      points(y1$t/7,y_scen1,type="l",col="black",lwd=3)
      points(y2$t/7,y_scen2,type="l",col="black",lwd=3)
      points(y4$t/7,y_scen4,type="l",col="black",lwd=3)
      points(y4$t/7,y_scen4a,type="l",col="black",lwd=3)
    }
    
  }
  
  ## pdf_fig_proof(file=paste0(tmpdir,"plot1.pdf"))
  plot_cum(-100,600,76,FALSE)
  ## dev.off()
  
  ## pdf_fig_proof(file=paste0(tmpdir,"plot2.pdf"))
  plot_cum(-100,4000,76,TRUE)
  ## dev.off()
  
  plot_cum(600)
  
  
  #' Load a database of social mixing patterns
  library(socialmixr)
  library(devtools)
  load_all()
  data(polymod)
  age_bounds <- c(0,5,11,18,30)
  nACs <- length(age_bounds)
  mixr_polyuk <- contact_matrix(
    polymod,
    countries="Great Britain",
    age.limits = age_bounds,
    symmetric=TRUE)

  #' Extract basic contact matrices
  matpolyuk <- mixr_polyuk$matrix
  poppolyuk <- mixr_polyuk$demography$population

  #' ## Define susceptibiity profiles and school closure mixing
  #'
  #' Establish some baseline values for susceptibility
  susBase <- rep(1,nACs)
  susLowKids <- susBase
  susLowKids[1:4] <- 0.333

  #' Run two test cases for the R0 parameterization
  yA <- cov_hybrid(
    vecN=poppolyuk,
    R0=1.01,
    matCt=matpolyuk,
    vecTcal=seq(0,20,0.001),
    vecInfNess=rep(1,nACs),
    vecSusTy=susLowKids,
    deterministic=TRUE)
  yB <- cov_hybrid(
    vecN=poppolyuk,
    R0=0.99,
    matCt=matpolyuk,
    vecTcal=seq(0,20,0.001),
    vecInfNess=rep(1,nACs),
    vecSusTy=susLowKids,
    deterministic=TRUE)

  plot(rowSums(yA$inf[,,1]),log="y",col="red",ylim=c(0.1,5))
  points(rowSums(yB$inf[,,1]),col="blue")

  #' Check to see if attack rates are about right with uniform mixing and
  #' uniform susceptibility and a perfectly balanced population. Seems to be OK
  #' to within half a percent.
  yC <- cov_hybrid(
    R0=1.8)
  a <- sum(yC$inf[,,1])/sum(yC$pop)
  epsilon <- (a + exp(-1.8*a) - 1)/a
  epsilon
}


```